# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data


The activity *.csv file is first read into R. The date column is set to the Date class in R. This enables the use of date functions to process the data.

### ANSWER 1 and 2 for this section

```{r}
 activity <- read.csv("activity.csv")
 activity$date <- as.Date(activity$date)
```

Options are set to ensure the display of real numbers is consistent.

```{r}
 options("scipen"=100, "digits"=4)
```


## What is the mean total number of steps taken per day?

### Analysis

Create a vector of the days that are part of the dataset. Using the "day" option in the seq() function creates a list of unique days in the date set. 

(The "day" parameter makes the sequence list each day only once while reading the ``` activity$date ``` column.)

```{r}
 daysequence <- seq(activity$date[1], activity$date[length(activity$date)],"day")
```

The resulting vector is displayed for completeness.

```{r}
 daysequence
```

There are **`r length(daysequence)`** total days in the dataset.

```{r}
length(daysequence)

```

There are **`r length(activity$date)/length(daysequence)`** 5-minute intervals in each day.

```{r}
length(activity$date)/length(daysequence)

```

This equals the number of 5-minute intervals in a 24 hour period.

```{r}
24*60/5

```

A numeric vector with length equal to the number of unique days is created to hold the total steps per day.

The ```for``` loop then adds up all steps for each day in the sequence.


```{r}

totalStepsPerDay <- numeric(length(daysequence))

for(i in 1:length(daysequence)){
     
     daystoadd <- activity$steps[(activity$date == daysequence[i])]
     
     totalStepsPerDay[i] <- sum(daystoadd, na.rm=TRUE)
          
 }
```

A barplot of total steps per day is created, by first setting up the labels for readability. Please note that each tick corresponds to a day in the ```daysequence``` vector displayed earlier. Note that this is for clarity, as will be described later. The barplot shows the steps for each day, where a histogram would show the *frequency* of a total steps count in the whole vector.


```{r}
Plotlabs <- seq(min(daysequence), max(daysequence), by="day")

Plotlabs.pos <- round(seq(1, length(daysequence), length.out=length(daysequence)))

mp <- barplot(totalStepsPerDay, xaxt="n", main="Plot of Total Steps Per Day in the dataset",
                xlab="Date", ylab="Number Of steps")

axis(1, at=mp, labels=Plotlabs)

```

**This barplot is shown to get a sense of the dataset, but was not required by the assignment.**

### ANSWER 1 for this section

This is the histogram required for the assignment. Note that it shows the frequencies for counts that fall in each of the "Sturges" bins. See ```?hist``` for details.


```{r}
hist(totalStepsPerDay, main="Histogram of Total Steps Per Day", xlab="Total count of steps per day")

```



### ANSWER 2 for this section

The **mean** and **median** of the total number of steps per day are calculated. 

The **mean** of total steps per day is `r mean(totalStepsPerDay)`, and the **median** is `r median(totalStepsPerDay)`.

```{r}
 mean(totalStepsPerDay)
``` 

```{r} 
 median(totalStepsPerDay)
``` 

The **summary** is also displayed for completeness.

```{r} 
 summary(totalStepsPerDay)
```

## What is the average daily activity pattern?

### Analysis

The ```aggregate``` function is used to rearrange the data into a data frame organized by *interval* and not by *date*. The first few items of the resulting data frame will be shown to validate this transformation.

```{r}
activity.DF <- data.frame(activity)

activity_mean_min <- aggregate(activity.DF[,1], by=list(activity.DF$date,activity.DF$interval), FUN=mean)
```

Reset the column names vector for clarity.

```{r}
colnames(activity_mean_min) <- c("Date", "Interval", "Steps")
```

Display the first few items in the data frame to show the effect.

```{r}
head(activity_mean_min)
```

Create vector to hold the sequence of intervals by subsetting by one date (the first day, but it would work with any day).

A numeric vector with length equal to the number of 5-minute intervals is then created to hold the mean steps per interval.

The ```for``` loop then takes the arithmetic mean for each interval in the sequence.

```{r}
intervalsequence <- activity$interval[(activity$date == daysequence[1])]

mean_steps_interval <- numeric(length(intervalsequence))

for(j in 1:length(intervalsequence)){
     
    intervaltoavg <- activity_mean_min$Steps[(activity_mean_min$Interval == intervalsequence[j])]
    
	  mean_steps_interval[j] <- mean(intervaltoavg, na.rm=TRUE)    
     
}
```

This is the plot of the average number of steps taken in a specific 5-minute interval, where the average is taken for that interval across all days in the dataset.


### ANSWER 1 for this section

```{r}
 plot(activity$interval[1:length(mean_steps_interval)], mean_steps_interval, 
      type="l", main="Plot of steps per 5-minute interval, averaged across all days", 
      xlab="5-minute Interval", 
      ylab="Average number of steps taken, averaged across all days")
```

*Note that the following plot is made by interval **index*** *and not by interval "identifier". The index ```1:288``` is continuous and will yield consistent plots very simply, using the "identifier" ```0, 5, ..., 2355``` can yield discontinuities in some plots unless labels and axes are modified. The plot above looks similar but has some discontinuities*

```{r}
 plot(1:length(mean_steps_interval), mean_steps_interval, 
      type="l", main="Plot of steps per 5-minute interval, averaged across all days", 
      xlab="5-minute Interval index", 
      ylab="Average number of steps taken, averaged across all days")
```

Both plots are shown because it not obvious from the problem statement which would be expected. Both seem acceptable, and the issues a factor of the identifiers used for the intervals, given that there **are** **288** 5-minute intervals in a 24 hour period, as shown previously.


The six number summary for the vector plotted above.

```{r}
 summary(mean_steps_interval)
``` 

This is the vector of average steps per interval used in the plot above.

```{r} 
 mean_steps_interval
``` 

The ```which.max``` function returns the index to the maximum value in the vector.
```{r} 
 which.max(mean_steps_interval)
``` 

This is the maximum average value per interval.
```{r} 
 mean_steps_interval[which.max(mean_steps_interval)]
``` 

This is the interval identifier for the maximum value.

### ANSWER 2 for this section

The 5-minute interval with the maximum steps per interval, averaged across all days in the dataset is **`r intervalsequence[which.max(mean_steps_interval)]`**.

```{r}
 intervalsequence[which.max(mean_steps_interval)]
```
 
### Validation

The interval "identifier" value, **`r intervalsequence[which.max(mean_steps_interval)]`**,  should match that above when subsetted from the original dataset using the index previously determined. Note that the index to the interval, **`r  which.max(mean_steps_interval)`**, can be visually verified from the printed ```mean_steps_interval``` vector.

```{r}
 activity[104, "interval"]
```

QED

## Imputing missing values

### Analysis

### ANSWER 1 for this section

There are **`r length(which(is.na(activity_mean_min)==TRUE))`** missing values in the dataset.

```{r}
 length(which(is.na(activity_mean_min)==TRUE))
```

### ANSWER 2 for this section

**The NA values per day per interval will be filled by the average number of steps across all days, for that interval, as follows:**

 - *Determine which days have NA step values*
 
 - *Insert the interval averages across all days vector into the steps for those days. (In effect making the average steps per interval the value for that interval.)*
 
 Assumptions:
 
 - All days have the same number of intervals.
 
 - All NA's ocurr in specific days.
 
 These assumptions were developed from initial exploration of the data set.
 
 One way to validate that the NA's ocurr on specific days is to take the mean of steps taken for each day in the ```daysequence```, removing the NA's in the mean function. This should result in ```NaN``` values where there are no values for steps.

```{r}
 
 mean_days <- numeric(length(daysequence))
 
 for(i in 1:length(daysequence)){

    daystoavg <- activity$steps[(activity$date == daysequence[i])]

	mean_days[i] <- mean(daystoavg, na.rm=TRUE)


 }

``` 

Display the vector ```mean_days``` to show ```NaN``` values.

```{r}

mean_days
```

We then subset ```daysequence``` to obtain a vector of days with NA's.

```{r}

NAdays <- daysequence[is.nan(mean_days)]

NAdays
```

### Validation

To check on this analysis, subset a vector from the previously generated result, and compare to expected values.

```{r}
 a <- activity_mean_min[(is.na(activity_mean_min$Steps)==TRUE), "Date"]
```

Length should equal the number of rows with NA's.

```{r}
length(a)
```

The number of unique should match the length of ```NAdays```.

```{r}
unique(a)
```

And the dates should be identical.

```{r}
identical(unique(a), NAdays)
```

QED

### ANSWER 3 in this section

Create a dataset with the NA's filled in. (As described in the process in answer 2 for this section).

```{r}

activity.DF <- data.frame(activity)

for(k in 1:length(NAdays)){
     
     activity.DF$steps[(activity.DF$date == NAdays[k])] <- mean_steps_interval
     
     
 }
 
```

Calculate the total steps per day as before, only now with the dataset with imputed values.

```{r}
 
 totalStepsPerDayIMP <- numeric(length(daysequence))
for(i in 1:length(daysequence)){
     
     daystoadd <- activity.DF$steps[(activity.DF$date == daysequence[i])]
     
     totalStepsPerDayIMP[i] <- sum(daystoadd, na.rm=TRUE)
     
     
 }
```


**This barplot is shown to get a sense of the dataset, but was not required by the assignment.**

```{r}

Plotlabs <- seq(min(daysequence), max(daysequence), by="day")

Plotlabs.pos <- round(seq(1, length(daysequence), length.out=length(daysequence)))

mp <- barplot(totalStepsPerDayIMP, xaxt="n", 
            main="Plot of Total Steps Per Day in the dataset, with Imputed values",
			xlab="Date", ylab="Number Of steps")
			
axis(1, at=mp, labels=Plotlabs)

```

### ANSWER 4 part a

This is the histogram required for the assignment. Note that it shows the frequencies for counts that fall in each of the "Sturges" bins. See ```?hist``` for details.

```{r}
hist(totalStepsPerDayIMP, main="Histogram of Total Steps Per Day, with Imputed Data",
     xlab="Total count of steps per day")

```

### ANSWER 4 parts b and c

The **mean** of total steps per day is `r mean(totalStepsPerDayIMP)`, and the **median** is `r median(totalStepsPerDayIMP)`.

```{r}
 mean(totalStepsPerDayIMP)
``` 
 
```{r} 
 median(totalStepsPerDayIMP)
```

The summary is shown for completeness.

```{r} 
 summary(totalStepsPerDayIMP)
```

### ANSWER 4 part d

*Do these values differ from the estimates from the first part of the assignment?*

- ANS: YES

The summaries for the two vectors, as well as the histograms, show the differences.

```{r} 
 summary(totalStepsPerDay)
```

```{r} 
 summary(totalStepsPerDayIMP)
```

*What is the impact of imputing missing data on the estimates of the total daily number of steps?*

 - ANS: The mean changes from `r mean(totalStepsPerDay)`, to `r mean(totalStepsPerDayIMP)` when imputed values are used. The median changes from `r median(totalStepsPerDay)` to `r median(totalStepsPerDayIMP)`. The histograms change shape, and the lower quantiles change. This points to a possible change in the density of the estimates. It is notable that the median equals the mean after imputation, when the mean was lower than the median before it.

## Are there differences in activity patterns between weekdays and weekends?

### ANSWER 1

Create a new factor variable in the dataset indicating whether a date is a weekend or a weekday. Weekend is defined by its dictionary definition of "Saturday and Sunday".

```{r}
 dayfactor <- character(length(activity$date))
 
 for(j in 1:length(activity.DF$date)){

  if((weekdays(activity.DF$date[j])=="Saturday")|(weekdays(activity.DF$date[j])=="Sunday"))
  	dayfactor[j] <- "weekend"
	else
		dayfactor[j] <- "weekday"
		
}	

 activity.DF <- cbind(activity.DF, dayfactor)
```



### Analysis

The ```aggregate``` function is used to rearrange the data into a data frame organized by *interval and dayfactor* and not by *date*. The first few items of the resulting data frame will be shown to validate this transformation.

```{r}

activity_mean_min_factor <- aggregate(activity.DF[,1], 
    by=list(activity.DF$interval,activity.DF$dayfactor), FUN=mean)
```

Reset the column names vector for clarity.

```{r}
colnames(activity_mean_min_factor) <- c("Interval", "DayFactor", "Steps")
```

Display the first few items in the data frame to show the effect.

```{r}

head(activity_mean_min_factor)
```


### ANSWER 2

Make a panel plot that contains a time series plot of steps taken on a 5-minute interval, averaged across weekend days and weekdays.

```{r}
 p <- xyplot(Steps ~ Interval | DayFactor,
             data = activity_mean_min_factor, 
             type = "l", main="Panel Time-Series Plot of steps 
             taken per 5-minute interval Weekend vs Weekday", 
             xlab="5-minute Interval", 
             ylab="Number of Steps averaged over Weekend or
             Weekday", layout=c(1,2,1))
 
 print(p)
```

*Note that the plot is made by interval **index*** *and not by interval "identifier". The index ```1:288``` is continuous and will yield consistent plots very simply, using the "identifier" ```0, 5, ..., 2355``` can yield discontinuities in some plots unless labels and axes are modified.*

```{r}

 intervalIndex_factor <- rep(1:length(intervalsequence), 2)

 activity_mean_min_factor <- cbind(activity_mean_min_factor, intervalIndex_factor)

 require(lattice)
 
 p <- xyplot(Steps ~ intervalIndex_factor | DayFactor,
              data = activity_mean_min_factor, type = "l", 
             main="Panel Time-Series Plot of steps taken 
             Weekend vs Weekday, per Interval Index", 
             xlab="Interval Index", 
             ylab="Number of Steps averaged over 
             Weekend or Weekday", layout=c(1,2,1))
 
 print(p)
 
```

The plot is drawn for the interval index as discussed above, to keep the code readable and accessible. The interval *identifiers* skip from 55 to 100, (i.e. ```...,55,100,...,155,200,...```), where the indices have the same increment from ```1:288```.

Both plots are shown because it not obvious from the problem statement which would be expected. Both seem acceptable, and the issues a factor of the identifiers used for the intervals, given that there **are** **288** 5-minute intervals in a 24 hour period, as shown previously.


